# Visualization 
EDA is a critical step in any data analysis workflow. After collecting and cleaning our WRDS dataset—focusing on public firms and removing outliers—we now explore key financial patterns. In this chapter, we:

Visualize industry-specific financial losses

Explore firm size and loss trends over time

Examine variable relationships via a correlation matrix

Evaluate earnings persistence using yearly regressions

We use high-quality, publication-ready graphics (PDF) and presentation-ready graphics (PNG) with R packages like ggplot2, corrplot, and modelsummary.


To learn more about the R packages used in this chapter, see the following tutorials:

[ggplot](https://ggplot2.tidyverse.org/) – for creating elegant data visualizations

[corrplot](https://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram?utm_source=chatgpt.com) – for visualizing correlation matrices

[modelsummary](https://modelsummary.com/.) – for formatting regression tables and model outputs


```{r}
#| label: load-libraries
#| echo: false
#| message: false
#| warning: false
#| results: hide

library(modelsummary)
library(kableExtra)
library(formattable)
library(lubridate)
library(glue)
library(haven)
library(fixest)
library(forcats)
library(tidyverse)
library(corrplot)  # if you're using correlation matrix
```




We work with a cleaned and winsorized version of firm-level financial data. The key variables include:

| Variable     | Description                                                  |
|--------------|--------------------------------------------------------------|
| `gvkey`      | Firm identifier                                              |
| `datadate`   | Date of observation                                          |
| `calyear`    | Calendar year                                                |
| `roa`        | Return on Assets                                             |
| `roa_lead_1` | Return on Assets in next year                                |
| `loss`       | Dummy: 1 if firm has a financial loss                        |
| `at`         | Total assets                                                 |
| `mve`        | Market value of equity                                       |
| `rd`         | R&D spending scaled by assets                                |
| `FF12`       | Industry name based on Fama-French 12 classification         |
| `ff12num`    | Industry code (numeric)                                      |



###6.2 Visualizing Losses by Industry
Objective:
Identify which industries are more prone to reporting losses, using FF12 categories.

Plot: Horizontal bar chart of % of firms with loss == 1 per FF12 category.

Interpretation:
This figure reveals financial vulnerability patterns. For instance, tech or healthcare might have higher losses due to R&D intensity, while utilities may show fewer.


```{r setup}

#| label: setup-path
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
data_path <- "E:/acct_995_data/abr"
```


```{r}
#| label: load-and-clean-data
#| echo: false
#| message: false
#| warning: false
#| results: hide

# Load helper scripts
source("src/-Global-Parameters.R")
source("src/utils.R")

# Read in the winsorized data
regdata <- read_dta(glue("{data_path}/regdata-R.dta")) |> 
  select(gvkey, datadate, calyear, roa, roa_lead_1, loss, at, mve, rd, FF12, ff12num)

```




```{r}

#| label: fig-losses-by-industry
#| fig-cap: "Proportion of Loss-Making Firms by Fama-French 12 Industry"
#| echo: false
#| message: false
#| warning: false


fig <- regdata |> 
  group_by(FF12) |> 
  summarize(pct_loss = sum(loss, na.rm = T)/n()) |> 
  #Next line reorders the FF12 industries to make them appear in order of %loss
  mutate(FF12 = forcats::fct_reorder(factor(FF12), (pct_loss))) |> 
  ggplot(aes(x = FF12, y= pct_loss)) + 
  geom_col(fill = "#0051ba") +
  # Fill color = Kansas Blue from : https://brand.ku.edu/guidelines/design/color
  scale_y_continuous(name = "Freq. of Losses", labels = scales::percent) +
  scale_x_discrete(name = "Fama-French Industry") +
  coord_flip() +
  #base_family = serif sets font to times new roman
  theme_bw(base_family = "serif") 

#Look at it in R  
fig

#For Latex output you might want to output to PDF
ggsave(glue("{data_path}/output/ff12_fig.pdf"), fig, width = 7, height = 6)

#For Word output you might want to output to an image such as .png
ggsave(glue("{data_path}/output/ff12_fig.png"), fig, width = 4.2, height = 3.6)

```
###6.3 Losses by Firm Size Over Time
Objective:
Compare financial stability across firms of different sizes over the years.

Plot:
Line chart of pct_loss over time, stratified by market value quintiles.

Interpretation:
Smaller firms often exhibit higher volatility and more losses. This dynamic changes across economic cycles

```{r}

#| label: fig-losses-by-size
#| fig-cap: "Proportion of Loss-Making Firms by Size Quintile Over Time"
#| echo: false
#| message: false
#| warning: false

fig <- regdata |> 
  group_by(calyear) |> 
  #create size quintiles by calyear
  mutate(size_qnt = factor(ntile(mve,5))) |> 
  group_by(calyear, size_qnt) |> 
  summarize(pct_loss = sum(loss, na.rm = T)/n()) |> 
  ggplot(aes(x = calyear, y= pct_loss, color = size_qnt, linetype = size_qnt)) + 
  geom_line() + geom_point() + 
  scale_y_continuous(name = "Freq. of Losses", labels = scales::percent) +
  scale_x_continuous(name = "Year", breaks = seq(1970,2025,5)) +
  #If you give these scales the same name they will appear in the same legend
  scale_color_discrete(name = "Size Quintile") +
  scale_linetype_discrete(name = "Size Quintile") +
  theme_bw(base_family = "serif") 

#Look at it in R  
fig

#For Latex
ggsave(glue("{data_path}/output/size_year.pdf"), fig, width = 7, height = 6)

#For Word
ggsave(glue("{data_path}/output/size_year.png"), fig, width = 7, height = 6)

```

###6.4 Correlation Matrix
Objective:
Understand relationships between core financial variables and detect multicollinearity.

Plot:
Heatmap-style correlation matrix (with coefficient values) for ROA, LOSS, R&D, etc.

Interpretation:
Strong correlation between roa and roa_lead_1 shows earnings persistence. Weak correlation between rd and roa may suggest R&D’s delayed or indirect impact.

```{r}


#| label: fig-correlation-matrix
#| fig-cap: "Correlation Matrix of Key Financial Variables"
#| echo: false
#| message: false
#| warning: false


library(corrplot)


corrdata <- regdata |> 
  select(`ROA_{t+1}` = roa_lead_1,
         `ROA_t` = roa, 
         `LOSS` = loss,
         `R\\&D` = rd,
         `TA` = at,
         `SIZE` = mve)

corrdata

correlation = cor(corrdata)
col2 = colorRampPalette(c('red', 'white', 'blue'))  

pdf(file=glue("{data_path}/output/corr_fig.pdf"))
corrplot(correlation, method = 'square', 
         addCoef.col = 'black', 
         diag = FALSE,
         tl.col='black', 
         type = 'full',
         tl.cex = 1,
         tl.srt = 0,
         tl.offset = 1,
         number.cex = 0.7,
         cl.ratio = 0.1,
         cl.pos = "r",
         col=col2(20),
         win.asp = .8)
dev.off()

```

###6.5 Yearly Earnings Persistence
Objective:
Test how well current ROA predicts future ROA, stratified by loss status.

Plot:
Line plot with confidence intervals from yearly regressions

Interpretation:
Firms without losses tend to show stronger earnings persistence. Loss firms show more variation and uncertainty in their earnings trajectory.



```{r}

#| label: fig-annual-regression-confidence-bands
#| fig-cap: "Annual Regressions of Future ROA on Current ROA with 95% Confidence Bands"
#| echo: false
#| message: false
#| warning: false


# bonus example 

figdata <- regdata |>
  #nest the regressions by year and loss
  nest_by(calyear,loss) |> 
  # fit the regressions
  mutate(
    fit = list(lm(roa_lead_1 ~ roa, data = data))
  ) |> 
  # use the broom package to tidy the regressions
  #option conf.int outputs the confidence intervals so we can plot them
  summarise(broom::tidy(fit, conf.int = TRUE)) |> 
  #I don't plan to plot the intercept so i will drop it from the data
  filter(term !="(Intercept)") 

#can also use this setup to do Fama-Macbeth regressions, etc. 
#can also use pmg package for Fama-Macbeth

#now make a ggplot object from the data
fig <- figdata |> 
  mutate(loss = factor(loss)) |> 
  ggplot(aes(x=calyear,y=estimate))  +
  geom_ribbon(aes(ymin = conf.low, ymax=conf.high, 
                  group = loss),
                  fill = "grey80") + 
  geom_line(aes(color=loss)) +
  geom_point(aes(color=loss)) +
  theme_bw(base_family = "serif") 

#Look at it in R  
fig

#For Latex
ggsave(glue("{data_path}/output/coef_year.pdf"), fig, width = 7, height = 6)

#For Word
ggsave(glue("{data_path}/output/coef_year.png"), fig, width = 7, height = 6)



```
